A Bidirected String Graph Model for Genome Assembly

### 2012-12-12 presentation/paper.tex
paper: updates

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
\usepackage{grffile}
%\usepackage{amsthm}

@@

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technologies can only
sequence short fragments of DNA.  As a result, the whole-genome shotgun
sequencing approach to genome sequencing relies on breaking up a sample of
genomic DNA into fragments, sequencing these fragments, then using a program
called a {\em genome assembler} to put together these short sequences to
reconstruct the original, much longer sequence.  This paper provides an
introduction and formalization of the genome assembly problem, then introduces a
model and algorithm for genome assembly that is based on a special type of graph
called a \KeyTerm{bidirected string graph}.  \end{abstract}

@@

@@

This definition does not account for the fact that the genomes of diploid and
polyploid organisms, including most plants and animals, contain multiple
slightly-differing copies of each chromosome.  In this case, the above
definition will consider each copy of each chromosome to be a completely
different string in the genome, unless the copies are completely identical.

@@

\begin{table}[H]
	\begin{center}
		\begin{tabular}{|c|>{\centering\arraybackslash}p{5.2cm}|}
			%\hline
			%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
			%\hline
			%Human mitochondrion &  $\sim$16,000 bp \\
			\hline
			{\bf Organism} & {\bf Approximate genome size} \\
			\hline
			{\it Variola} virus (smallpox) &  186,000 bp\\
			\hline
			{\it E. coli} (a bacterium) & 4,600,000 bp \\
			\hline
			{\it Ananas comosus} (domesticated pineapple) & 500,000,000 bp \\
			\hline
			{\it Zea mays} (domesticated corn) & 2,000,000,000 bp \\
			\hline
			{\it Homo sapiens} (human)       &  3,200,000,000 bp \\
			\hline
		\end{tabular}
	\end{center}
	\caption{Approximate genome sizes of selected organisms}
	\label{tab:GenomeSizes}
\end{table}

@@

\KeyTerm{DNA sequencing} is the process of determining the base pair sequence of
(\KeyTerm{sequencing}) one or more DNA molecules.  \KeyTerm{Genome sequencing}
is the use of DNA sequencing to determine the base pair sequence of an entire
genome.

@@

Therefore, to sequence an entire genome, which may be millions of base pairs
long, the DNA from the whole genome must be broken into much smaller fragments,
which are then sequenced.\footnote{This specifically refers to the
\KeyTerm{whole-genome shotgun sequencing} approach to genome assembly rather
than alternatives such as \KeyTerm{BAC-to-BAC} sequencing that are able to
restrict sequences to certain regions of the genome in order to make sequencing
the genome easier (currently, at the disadvantage of increased cost and
non-computational work).} Following this, the original genome must be
computationally reconstructed by a computer program called a \KeyTerm{genome
assembler} in a process called \KeyTerm{genome assembly}.  \KeyTerm{Genome
sequencing} therefore refers to the entire process of sequencing a genome,
including both lab work and computational work, while \KeyTerm{genome assembly}
only refers to the computational work.\footnote{It usually also takes some
technology-specific computational work to actually prepare the sequences for
assembly; for example, a series of digital photographs of clumps of
fluorescently labeled molecules might need to be interpreted as reads using
image-processing algorithms.  However, these programs and/or algorithms
generally have already been developed by the manufacturer of the sequencing
machine and work fairly well, and ``genome assembly'' does not cover these
algorithms, although assemblers still may be tuned for specific sequencing
technologies.}

@@

The goal of genome assembly is to reconstruct a genome, as defined earlier,
given a set of substrings that may originate from any strand of any DNA molecule
in the genome.  Each substring is called a \KeyTerm{read} and is the sequence of
some DNA that was sequenced by the sequencing machine.  In general, reads may
contain errors, including \KeyTerm{substitutions}, where a base pair in the read
was incorrectly read as some other base pair, as well as \KeyTerm{insertions},
where a sequence of one or more random base pairs have been inserted into some
position in the read, and \KeyTerm{deletions}, where a sequence of one or more
base pairs from the original genome is missing from the interior of the read.
Reads are expected to be uniformly distributed on the genome with a certain
coverage $X$ depending on how much data was collected, meaning that each base
pair in the original genome is expected to be included in $X$ different reads on
average, allowing for the reads to be from either strand.  However, some
sequencing technologies produce biased data where the coverage across the genome
is much more variable than that expected by uniform, random sampling.

@@

It is also possible for reads to come from foreign DNA; for example, if a DNA
sample is obtained from grinding up plant or animal tissue, there could be some
bacterial cells, each with its own genome, contaminating the sample.  Ideally,
an assembler should be prepared to completely ignore a read if it seems too
unusual.  However, this problem is not considered in this paper.

>>>>

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
\usepackage{grffile}
\usepackage{threeparttable}
%\usepackage{hyperref}
%\usepackage{amsthm}

@@

Sequencing an organism's genome is an important step to further genetic research
on that organism.  However, currently available sequencing technologies can only
sequence short fragments of DNA.  As a result, the whole-genome shotgun
sequencing approach to genome sequencing relies on breaking up a sample of
genomic DNA into short fragments, sequencing these fragments, then using a
program called a {\em genome assembler} to put together these short sequences to
reconstruct the original genome.  This paper provides an introduction and
formalization of the genome assembly problem, then introduces a model and
algorithm for genome assembly that is based on a special type of graph called a
\KeyTerm{bidirected string graph}.  \end{abstract}

@@

\label{sec:genome}

@@

This definition does not account for the fact that the genomes of diploid and
polyploid organisms, including most plants and animals, contain multiple
slightly-differing copies of each chromosome.  In this case, this definition
will consider each copy of each chromosome to be a completely different string
in the genome, unless the copies are completely identical.

@@

\begin{table}[H]
	\begin{center}
	\caption{Approximate genome sizes of selected organisms}
		\label{tab:GenomeSizes}
		\begin{threeparttable}
			\begin{tabular}{|l|>{\arraybackslash}p{5.2cm}|}
				%\hline
				%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
				%\hline
				%Human mitochondrion &  $\sim$16,000 bp \\
				\hline
				{\bf Organism} & {\bf Approximate genome size} \\
				\hline
				{\it Variola} virus (smallpox) \tnote{1} &  186,000 bp \cite{smallpox}\\
				\hline
				{\it E. coli} (a bacterium) & 4,600,000 bp \\
				\hline
				{\it Ananas comosus} (domesticated pineapple) & 500,000,000 bp \\
				\hline
				{\it Zea mays} (domesticated corn) & 2,400,000,000 bp \cite{maize} \\
				\hline
				{\it Homo sapiens} (human)       &  3,200,000,000 bp \\
				\hline
			\end{tabular}
			\begin{tablenotes}
				\item [1] Technically, viruses are not organisms, but they still
				have genomes.  Actually, some viruses have genomes made of RNA,
				not DNA, but this is not considered in this paper.
			\end{tablenotes}
		\end{threeparttable}
	\end{center}
\end{table}

@@

\KeyTerm{DNA sequencing} is the process of determining the base pair sequence of
(\KeyTerm{sequencing}) one or more DNA molecules.  \KeyTerm{Genome sequencing}
is the process of sequencing of all the DNA molecules that make up a genome.

@@

Therefore, to sequence an entire genome, which may be millions of base pairs
long, the DNA from the whole genome must be broken into much smaller fragments,
which are then sequenced.\footnote{This specifically refers to the
\KeyTerm{whole-genome shotgun sequencing} approach to genome assembly rather
than alternatives such as \KeyTerm{BAC-to-BAC} sequencing that are able to
restrict sequences to certain regions of the genome in order to make sequencing
the genome easier (currently, at the disadvantage of increased cost and much more
non-computational work).} Following this, the original genome must be
computationally reconstructed by a computer program called a \KeyTerm{genome
assembler} in a process called \KeyTerm{genome assembly}.  \KeyTerm{Genome
sequencing} therefore refers to the entire process of sequencing a genome,
including both lab work and computational work, while \KeyTerm{genome assembly}
only refers to the computational work.\footnote{It usually also takes some
technology-specific computational work to actually prepare the sequences for
assembly; for example, a series of digital photographs of clumps of
fluorescently labeled molecules might need to be interpreted as reads using
image-processing algorithms.  However, these programs and/or algorithms
generally have already been developed by the manufacturer of the sequencing
machine and are assumed to work fairly well, so ``genome assembly'' does not cover these
algorithms.  Still, assemblers may be tuned for specific sequencing
technologies.}

@@

The goal of genome assembly is to reconstruct a genome, as defined in
\Section{sec:genome}, given a set of substrings that may originate from any
strand of any DNA molecule in the genome.  Each substring is called a
\KeyTerm{read} and is the sequence of some DNA that was sequenced by the
sequencing machine.  In general, reads may contain errors, including
\KeyTerm{substitutions}, where a base pair in the read was incorrectly read as
some other base pair, as well as \KeyTerm{insertions}, where a sequence of one
or more random base pairs have been inserted into some position in the read, and
\KeyTerm{deletions}, where a sequence of one or more base pairs from the
original genome is missing from the interior of the read.  Reads are expected to
be uniformly distributed on the genome with a certain coverage $X$ depending on
how much data was collected, meaning that each base pair in the original genome
is expected to be included in $X$ different reads on average, allowing for the
reads to be from either strand.  However, some sequencing technologies produce
biased data where the coverage across the genome is much more variable than that
expected by uniform, random sampling.

@@

It is also possible for reads to come from foreign DNA; for example, if a DNA
sample is obtained by grinding up plant or animal tissue, there could be some
bacterial cells, each with its own genome, contaminating the sample.  Ideally,
an assembler should be prepared to completely ignore a read if it seems too
unusual.  However, this problem is not considered in this paper.
